Morus: Biases in the stream

### 2018-05-04 morusAC.tex
smalltypos

{\tt StepUpdate} is composed of 5 rounds with similar operations to update
the state $S^t$, in which the additional input $m^t$ is used in rounds 2 to 5, but not in round 1. Each round uses the word-wise left circular shift, which divides a $q$-bit register value into 4 words of $q/4$ bits, and performs left circular shift operation for every $q/4$-bit word. Throughout the paper, we denote the word size by $w$, i.e. $w=32$ for \cipher{MORUS640} and $w=64$ for \cipher{MORUS1280}. The word-wise circular shift constants $b_i$ in round $i$ are defined in Table~\ref{Tbl:rcon}. Besides, the left rotation of a whole $q$-bit register is used. The register-wise circular shift constants are also listed in Table~\ref{Tbl:rcon}.
\begin{table}[!htb]
\centering
\caption{Left Circular-Shift Constants of MORUS} \label{Tbl:rcon}
\begin{tabular}{c||ccccc|ccccc} \hline
& \multicolumn{5}{c|}{word-wise circular shift} & \multicolumn{5}{c}{register-wise circular shift} \\
                  & \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
\cipher{MORUS640} &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
\cipher{MORUS1280}& 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\ \hline
\end{tabular}
\end{table}
We denote those register-wise and word-wise left circular shift by $x \rotl y$ and $x \rotlxy y$, respectively. (Those are denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \cipher{MORUS} design document.)

@@

\subsubsection{Initialization.}
The initialization of \cipher{MORUS640} consists of loading a 128-bit key $K$ and a 128-bit nonce IV into the state, and running the {\tt StateUpdate} for 16 steps. The key and IV are loaded into the state as follows:
\begin{align*}
S^{-16}_0 = \textrm{IV}_{128}, &&
S^{-16}_1 = K_{128}, &&
S^{-16}_2 = 1^{128}, &&
S^{-16}_3 = const_0, &&
S^{-16}_4 = const_1.
\end{align*}
Then the internal state is updated 16 times by  ${\tt StateUpdate}(S^t,0)$ for $t=-16,-15,\cdots,-1$. Finally, the key is xored to the state again as $S^0_1 \leftarrow S^0_1 \oplus K$.

@@

\subsubsection{Associated Data Processing.}
After the initialization, the associated data $A$ is processed. For the padding, if the last associated data block is not a full block, use `0' bits to pad it to 128 bits and 256 bits for \cipher{MORUS640} and \cipher{MORUS1280}, respectively, and the padded full block is used to update the state. Note that if the length of $A$ dented by `$adlen$' is 0, the associated data processing is skipped.

Then, state is updated by $S^{t+1} \leftarrow {\tt StateUpdate}(S^t, A^t)$ for $t=0,1,\cdots,\ell$ where $\ell = \lceil \frac{adlen}{128} \rceil -1$ and $\ell = \lceil \frac{adlen}{256} \rceil -1$ for \cipher{MORUS640} and \cipher{MORUS1280}, respectively.

@@

In \cipher{MORUS640}, Let $u = \lceil \frac{adlen}{128} \rceil$ and $v = \lceil \frac{msglen}{128} \rceil$. The following is performed for $t=0, 1, \cdots, v-1$;
\begin{align*}
C^t &\leftarrow M^t \oplus S^{u+t}_0 \oplus (S^{u+t}_1 \lll 96) \oplus (S^{u+t}_2 \cdot S^{u+t}_3),\\
S^{u+t+1} &\leftarrow {\tt StateUpdate}(S^{u+t},M^t).
\end{align*}

@@

\subsubsection{Finalization.}
The finalization generates the authentication tag $T$ using 10 more steps. We only discuss the case that $T$ is not truncated. The associated data length and the message length are used to update the state as follows.
\begin{enumerate}
\item $tmp \leftarrow adlen\|msglen$ for \cipher{MORUS640} and $tmp \leftarrow adlen\|msglen\|0^{128}$ for \cipher{MORUS1280}, where $adlen$ and $msglen$ are represented as 64-bit integers.
\item $S^{u+v}_4 \leftarrow S^{u+v}_4 \oplus S^{u+v}_0.$
\item For $t = u+v, u+v+1, \cdots, u+v+9,$ compute
%\begin{equation*}
$S^{t+1} \leftarrow {\tt StateUpdate} (S^t, tmp)$.
%\end{equation*}
\item $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 96) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ for \cipher{MORUS640} and $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 192) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ for \cipher{MORUS1280}.
\end{enumerate}

@@

We begin with a few observations about the \StateUpdate{} function. Beside XOR and AND operations, the \StateUpdate{} function uses two types of bit rotations:
\begin{enumerate}
\item \emph{word-wise} rotations perform a circular shift on each word within a register;
\item \emph{register-wise} rotations perform a circular shift on a whole register.
\end{enumerate}
The second type of rotation always shifts registers by a multiple of the word size $w$. This amounts to a (circular) permutation of the words within the register: for example, if a register contains the words $(A, B, C, D)$, and a register-wise rotation by $w$ bits to the left is performed, then the register now contains the words $(B, C, D, A)$.

>>>>

{\tt StepUpdate} is composed of 5 rounds with similar operations to update
the state $S^t$, in which the additional input $m^t$ is used in rounds 2 to 5, but not in round 1. Each round uses the word-wise left circular shift, which divides a $q$-bit register value into 4 words of $q/4$ bits, and performs left circular shift operation for every $q/4$-bit word. Throughout the paper, we denote the word size by $w$, i.e. $w=32$ for \cipher{MORUS640} and $w=64$ for \cipher{MORUS1280}. The word-wise circular shift constants $b_i$ in round $i$ are defined in Table~\ref{Tbl:rcon}. Besides, the left rotation of a whole $q$-bit register is used. The register-wise circular shift constants are also listed in Table~\ref{Tbl:rcon}.
\begin{table}[!htb]
\centering
\caption{Left Circular-Shift Constants of MORUS} \label{Tbl:rcon}
\begin{tabular}{c||ccccc|ccccc} \hline
& \multicolumn{5}{c|}{word-wise circular shift} & \multicolumn{5}{c}{register-wise circular shift} \\
                  & \makebox[2em]{$b_0$} & \makebox[2em]{$b_1$} & \makebox[2em]{$b_2$} & \makebox[2em]{$b_3$} & \makebox[2em]{$b_4$} & \makebox[2em]{$b'_0$} & \makebox[2em]{$b'_1$} & \makebox[2em]{$b'_2$} & \makebox[2em]{$b'_3$} & \makebox[2em]{$b'_4$} \\ \hline
\cipher{MORUS640} &  5 & 31 &  7 & 22 & 13 & 32 &  64 &  96 &  64 & 32 \\
\cipher{MORUS1280}& 13 & 46 & 38 &  7 &  4 & 64 & 128 & 192 & 128 & 64 \\ \hline
\end{tabular}
\end{table}
We denote those register-wise and word-wise left circular shift by $x \lll y$ and $x \lll_w y$, respectively. (Those are denoted by \texttt{<<<} and \texttt{Rotl\_xxx\_yy} respectively in the \cipher{MORUS} design document.)

@@

\subsubsection{Initialization.}
The initialization of \cipher{MORUS640} consists of loading a 128-bit key $K$ and a 128-bit nonce IV into the state, and running the {\tt StateUpdate} for 16 steps. The key and IV are loaded into the state as follows:
\begin{align*}
S^{-16}_0 = \textrm{IV}_{128}, &&
S^{-16}_1 = K_{128}, &&
S^{-16}_2 = 1^{128}, &&
S^{-16}_3 = const_0, &&
S^{-16}_4 = const_1.
\end{align*}
Then the internal state is updated 16 times by  ${\tt StateUpdate}(S^t,0)$ for $t=-16,-15,\ldots,-1$. Finally, the key is xored to the state again as $S^0_1 \leftarrow S^0_1 \oplus K$.

@@

\subsubsection{Associated Data Processing.}
After the initialization, the associated data $A$ is processed. For the padding, if the last associated data block is not a full block, use `0' bits to pad it to 128 bits and 256 bits for \cipher{MORUS640} and \cipher{MORUS1280}, respectively, and the padded full block is used to update the state. Note that if the length of $A$ denoted by `$adlen$' is 0, the associated data processing is skipped.

Then, state is updated by $S^{t+1} \leftarrow {\tt StateUpdate}(S^t, A^t)$ for $t=0,1,\ldots,\ell$ where $\ell = \lceil \frac{adlen}{128} \rceil -1$ and $\ell = \lceil \frac{adlen}{256} \rceil -1$ for \cipher{MORUS640} and \cipher{MORUS1280}, respectively.

@@

In \cipher{MORUS640}, Let $u = \lceil \frac{adlen}{128} \rceil$ and $v = \lceil \frac{msglen}{128} \rceil$. The following is performed for $t=0, 1, \ldots, v-1$;
\begin{align*}
C^t &\leftarrow M^t \oplus S^{u+t}_0 \oplus (S^{u+t}_1 \lll 96) \oplus (S^{u+t}_2 \cdot S^{u+t}_3),\\
S^{u+t+1} &\leftarrow {\tt StateUpdate}(S^{u+t},M^t).
\end{align*}

@@

\subsubsection{Finalization.}
The finalization generates the authentication tag $T$ using 10 more steps. We only discuss the case that $T$ is not truncated. The associated data length and the message length are used to update the state as follows.
\begin{enumerate}
\item $tmp \leftarrow adlen\|msglen$ for \cipher{MORUS640} and $tmp \leftarrow adlen\|msglen\|0^{128}$ for \cipher{MORUS1280}, where $adlen$ and $msglen$ are represented as 64-bit integers.
\item $S^{u+v}_4 \leftarrow S^{u+v}_4 \oplus S^{u+v}_0.$
\item For $t = u+v, u+v+1, \ldots, u+v+9,$ compute
%\begin{equation*}
$S^{t+1} \leftarrow {\tt StateUpdate} (S^t, tmp)$.
%\end{equation*}
\item $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 96) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ for \cipher{MORUS640} and $T = S^{u+v+10}_0 \oplus (S^{u+v+10}_1 \lll 192) \oplus ( S^{u+v+10}_2 \cdot S^{u+v+10}_3)$ for \cipher{MORUS1280}.
\end{enumerate}

@@

We begin with a few observations about the \StateUpdate{} function. Besides XOR and AND operations, the \StateUpdate{} function uses two types of bit rotations:
\begin{enumerate}
\item \emph{word-wise} rotations perform a circular shift on each word within a register;
\item \emph{register-wise} rotations perform a circular shift on a whole register.
\end{enumerate}
The second type of rotation always shifts registers by a multiple of the word size $w$. This amounts to a (circular) permutation of the words within the register: for example, if a register contains the words $(A, B, C, D)$, and a register-wise rotation by $w$ bits to the left is performed, then the register now contains the words $(B, C, D, A)$.
